---
title: "Class8"
author: "David Majeed (A17885958)"
format: pdf
toc: true
editor: visual
---

## Intro

Our goal is to analysis data coming from breast cancer research using the techniques we have learnt in other classes.

## Data Import

The data comes from fine needle aspiration

```{r}

library(readr)


fna.data <- read_csv("Class8/WisconsinCancer.csv")
#read.csv allows us to import our data set

wisc.df<- read.csv("Class8/WisconsinCancer.csv", row.names=1)
#Shift the row names by 1 to line up our data properly

head(wisc.df,3)
#Not to show up all the data

#lets remove the diagonsis column
wisc.data<- wisc.df[,-1]
#make it's own
diagonsis<- wisc.df[,1]
```

## Exploring the Data

Q1. There are 569 observation

```{r}
nrow(wisc.data)
```

Q2. There are 212 malignant diagnoses

```{r}
#my attempt
sum(diagonsis=='M')
#Prof's way
table(diagonsis)
```

Q3. There are 10 variables in the data with the suffix of \_mean

```{r}
length(grep("_mean", colnames(wisc.data), value=F))
#length counts
#grep searches for the suffix
```

## PCA

Q4. The orginal variance is 0.4427 from PC1

Q5. PC3 is required to desrcibe at least 70% of the original variance

Q6. PC7 is needed for at least 90% of the original variance data

```{r}
#we need to scale our data
wisc.pr<-prcomp(wisc.data, scale=T)
summary(wisc.pr)
```

Q7. The thing that stands out is the lack of readability, all the labels and data points are stacked one another thus I can not understand it at all

```{r}
biplot(wisc.pr)

```

```{r}
#lets make a PC score plot
library(ggplot2)
ggplot(wisc.pr$x)+ 
  aes(PC1, PC2, col=diagonsis)+
  geom_point()
```

Q8. I notice that it has shifted everything down a bit. It still looks pretty similar to the previous plot but with more mixing of the two types together.

```{r}
ggplot(wisc.pr$x)+ 
  aes(PC1, PC3, col=diagonsis)+
  geom_point()
```

## Variance

```{r}
pr.var<- wisc.pr$sdev^2
head(pr.var)
```

\\\\\\\\

\\\\\\

\\\\\\

\\\\\\

\\\\\\

\\\\\\

## Hierarchical Clustering

\\\\\\

\\\\\\\\

\\\\\\\\

\\\\\\\\

## Combining methods

```{r}
d<-dist(wisc.pr$x[,1:4])
wisc.pr.hclust<-hclust(d, method="ward.D2")
plot(wisc.pr.hclust)
abline(h=70, col="red")

grps<-cutree(wisc.pr.hclust, h=70)
table(grps)
```

How does this clustering pattern correspond to expert diagnosis

```{r}
table(diagonsis, grps)
```
