---
title: "Class 7"
author: "David Majeed (A17885958)"
format: pdf
toc: true
editor: visual
---

## Background

Today we will explore some core machine learning methods that are very popular.

These include **clustering** and **dimensional reduction** K-Means Clustering

The main function for K-means clustering is `kmeans()` Lets make some simple data with the `rnorm()`

```{r}
x<- c(rnorm(30, -3),
rnorm(30, 3))
x

#c() makes a vector, rnorm generates numbers from a normal distribution

z<- cbind(x=x, y=rev(x))
plot(z)

#cbind puts the two data sets of x and rev(x) together


```

```{r}
km<- kmeans(z, 2)
#we can now use kmeans, the first input is z which is our data it will
#look at. The 2 is the number of clusters
#We can dive deeper with the attributes() function
attributes(km)

#Q? How many points in each cluster?
sum(km$cluster==2)
sum(km$cluster==1)
#Or use this
km$size

#Q? What componet of your result object 
#details cluster assignmnet/membership
km$cluster

#Q? Can we plot our data 'z' and color by kmeans cluster 
#and add cluster centers as blue points

plot(z,col=km$cluster)
points(km$centers, col="blue", pch=15)

#points() put points on the graph and pch=makes the points bigger

#Q? Run a K-means clustering and plot the results asking for 4 cluster

kn<-kmeans(z, 4)
kn
plot(z,col=kn$cluster)
points(kn$centers, col="blue", pch=15)
```

One approach is to try different values for center and then pick the best

```{r}
ans<- NULL
for(i in 1:10){
km<-kmeans(z,i)

ans<- c(ans, km$tot.withinss)

}
plot(ans, type="o",
     xlab="Number of clusters",
     ylab= "Total Sum of Squares")
```

Hierarchical Clustering

The main function for hierarchical clustering is `hclust()`

The function does not take your "raw" data, you need a distance matrix

```{r}
d<-dist(z)
hc<-hclust(d)
hc

```

There is a bespoke plot() method for hclust() result objects

```{r}
plot(hc)
abline(h=8, col="red")
#This is how we see the cross bars, the abline() draws a line
#We can cut it to reveal the cluster

grps <- cutree(hc, k=2)

#Q? Make a plot of 'z' with yoyr hclust results (color by cluster membership)

plot(z, col=grps)
```

## Principal Component Analysis (PCA)

PCA is a dimensional reduction method that is popular for revealing patterns in complex datasets

Analysis of UK Food

```{r}

url <- "https://tinyurl.com/UK-foods" 
x <- read.csv(url) 
#load our data from the internet

nrow(x) 
ncol(x) 
dim(x) 
#lets find how many rows and columns

```

Q1. 17 rows and 5 columns

```{r}
View(x)
#lets us view the code

#lets set up the rows properly
url <- "https://tinyurl.com/UK-foods" 
x <- read.csv(url)
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
View(x)
dim(x)
#restate x in order to define it properly for these steps
```

```{r}
x <- read.csv(url, row.names=1)
head(x)
#Alt. way to what we did above
```

Q.2 I prefer the second method as it is more efficient to code needing less lines and thus leading to less room for mistakes. I was stuck on the first method due to an error in the code which the second method was able to handle much faster. In the circumstance that you do not know how the table looks, the second method is better as it directly goes to the .cvs file.

## Bar Plot

```{r}
#lets graph it
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))

#If we want to stack it we can remove the besides=T
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))

```

Q.3 By turning the besides = false we can stack it

```{r}
#install.packages("tidyr")
library(tidyr)

# Convert data to long format for ggplot with `pivot_longer()`
x_long <- x |> 
          tibble::rownames_to_column("Food") |> 
          pivot_longer(cols = -Food, 
                       names_to = "Country", 
                       values_to = "Consumption")

dim(x_long)

#Lets make ggplot for a grouped bar plot
library(ggplot2)

ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "dodge") +
  theme_bw()

#Lets stack it
ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col() +
  theme_bw()
```

Q.4 Similarly to the barplot function, we just need to remove a piece of the code to stack it. For ggplot() we remove "position ="dodge"" which then will stack it

## Pairwise & Heat Map

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)

```

Q.5 Let's make a pairwise plot. This is done using the pairs() function, we then use 'x' to signfiy which data set, and then, we use col=rainbow(nrow(x)) to color the dots based on the row they come from, and finally, pch=16 allows us to control the size of the dots. The graph then maps each country to one another and compares them, if they all line up on a diagonal then the data points are similar to one another

```{r}
#lets make some heat maps
#install.packages("pheatmap")
library(pheatmap)

pheatmap( as.matrix(x) )

```

Q.6 From the heatmap, we can see that Wales and England are most similar to one another. The biggest difference between N. Ireland and the rest of the UK is in alcoholic drinks

The main function for PCA is prcomp(). The function wants the obesrevations to be the rows and the variables to be the columns

## PCA Plots

Q.7

```{r}
pca <- prcomp (t(x))

# Create a data frame for plotting
df <- as.data.frame(pca$x)
df$Country <- rownames(df)

# Plot PC1 vs PC2 with ggplot
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()
```

Q.8

```{r}
t(x)
#transposes our data
pca <- prcomp (t(x))
summary(pca)

#The returned pca objects has components that we can use to make our main result figures

attributes(pca)

#The main results figure from this analysis is called a PC scored plot

pca$x

my_color<- c("orange","red","blue", "green")

library(ggplot2)
ggplot(pca$x)+
  aes(PC1,PC2, label=row.names(pca$x))+
  geom_point(col=my_color)+
  geom_text(size=3,vjust=2, col=my_color)


```

The PC score plot allows us to visually see how the countries vary from one another in a 2d space despite representing 17 data axis. From it we can see that Northern Ireland is the most different than any other country in the United Kingdom.

```{r}
ggplot(pca$rotation)+
  aes(PC1, reorder(row.names(pca$rotation),PC1))+
  geom_col()
```

The loading plot allows us to see how each of the different categories differ from one another, this is in contrast to the last plot where we were able to see differences among countries but not the categories. This plot can help us find which data axis we should investigate more into.

Q.9. The two food groups are fresh potatoes and soft drinks. PC2 is the axis that spans the second most variation after

```{r}
ggplot(pca$rotation) +
  aes(x = PC2, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC2 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```
